---
title: "Introduction to linear regression"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## Batter up 

The movie [Moneyball](http://en.wikipedia.org/wiki/Moneyball_(film)) focuses on
the "quest for the secret of success in baseball". It follows a low-budget team, 
the Oakland Athletics, who believed that underused statistics, such as a player's 
ability to get on base, better predict the ability to score runs than typical 
statistics like home runs, RBIs (runs batted in), and batting average. Obtaining 
players who excelled in these underused statistics turned out to be much more 
affordable for the team.

In this lab we'll be looking at data from all 30 Major League Baseball teams and
examining the linear relationship between runs scored in a season and a number 
of other player statistics. Our aim will be to summarize these relationships 
both graphically and numerically in order to find which variable, if any, helps 
us best predict a team's runs scored in a season.

## The data

Let's load up the data for the 2011 season.

```{r load-data, eval=TRUE}
load("more/mlb11.RData")
library(ggplot2)
library(magrittr)
```

In addition to runs scored, there are seven traditionally used variables in the 
data set: at-bats, hits, home runs, batting average, strikeouts, stolen bases, 
and wins. There are also three newer variables: on-base percentage, slugging 
percentage, and on-base plus slugging. For the first portion of the analysis 
we'll consider the seven traditional variables. At the end of the lab, you'll 
work with the newer variables on your own.

1.  What type of plot would you use to display the relationship between `runs` 
    and one of the other numerical variables? Plot this relationship using the 
    variable `at_bats` as the predictor. Does the relationship look linear? If 
    you knew a team's `at_bats`, would you be comfortable using a linear model 
    to predict the number of runs?

**We can use a scatter plot to show the relationship between `runs` and `at_bats`. The linear fit model below does appear to indicate a linear relationship between the independent variable `at_bats` and the response variable `runs`.**

```{r,fig.height=3}
ggplot(mlb11, aes(x=runs, y=at_bats)) + geom_point() + geom_smooth(method='lm')
```


If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor, eval=TRUE}
cor(mlb11$runs, mlb11$at_bats)
```

## Sum of squared residuals

Think back to the way that we described the distribution of a single variable. 
Recall that we discussed characteristics such as center, spread, and shape. It's
also useful to be able to describe the relationship of two numerical variables, 
such as `runs` and `at_bats` above.

2.  Looking at your plot from the previous exercise, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.

**The relationship between the two variables appears to be linear and directly proportional. The slope indicates that exists a strong positive correlation between `runs` and `at_bats` . Also, there is a high degree of variability in the scatter plot. We could more easily see this variability in a residuals plot from the fit line.**



Just as we used the mean and standard deviation to summarize a single variable, 
we can summarize the relationship between these two variables by finding the 
line that best follows their association. Use the following interactive 
function to select the line that you think does the best job of going through 
the cloud of points.

```{r plotss-atbats-runs, eval=TRUE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs)
```

After running this command, you'll be prompted to click two points on the plot 
to define a line. Once you've done that, the line you specified will be shown in
black and the residuals in blue. Note that there are 30 residuals, one for each 
of the 30 observations. Recall that the residuals are the difference between the
observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes
the sum of squared residuals. To visualize the squared residuals, you can rerun 
the plot command and add the argument `showSquares = TRUE`.

```{r plotss-atbats-runs-squares, eval=TRUE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the slope and
intercept of your line as well as the sum of squares.

3.  Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?

**The interactivity of the function did not work. I could see briefly a screen flashing "click two points" but it dissapeared quicly to a default plot of residuals.**
    
```{r plotss-atbats-runs2, eval=TRUE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs)
```

## The linear model

It is rather cumbersome to try to get the correct least squares line, i.e. the 
line that minimizes the sum of squared residuals, through trial and error. 
Instead we can use the `lm` function in R to fit the linear model (a.k.a. 
regression line).

```{r m1, eval=TRUE}
m1 <- lm(runs ~ at_bats, data = mlb11)
```

The first argument in the function `lm` is a formula that takes the form 
`y ~ x`. Here it can be read that we want to make a linear model of `runs` as a 
function of `at_bats`. The second argument specifies that R should look in the 
`mlb11` data frame to find the `runs` and `at_bats` variables.

The output of `lm` is an object that contains all of the information we need 
about the linear model that was just fit. We can access this information using 
the summary function.

```{r summary-m1, eval=TRUE}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe 
the model is shown at the top. After the formula you find the five-number 
summary of the residuals. The "Coefficients" table shown next is key; its first 
column displays the linear model's y-intercept and the coefficient of `at_bats`.
With this table, we can write down the least squares regression line for the 
linear model:

\[
  \hat{y} = -2789.2429 + 0.6305 * atbats
\]



One last piece of information we will discuss from the summary output is the 
Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the 
proportion of variability in the response variable that is explained by the 
explanatory variable. For this model, 37.3% of the variability in runs is 
explained by at-bats.

4.  Fit a new model that uses `homeruns` to predict `runs`. Using the estimates 
    from the R output, write the equation of the regression line. What does the 
    slope tell us in the context of the relationship between success of a team 
    and its home runs?

```{r m2, eval=TRUE}
m2 <- lm(runs ~ homeruns, data = mlb11)
summary(m2)
```

\[
  \hat{y} = 415.2389 + 1.8345 * atbats
\]

**Obviously, there is a strong and positive relationship between the homeruns of a teams and the number of runs of a teams. Homeruns lead to victory.**


```{r}
ggplot(mlb11, aes(x=runs, y=homeruns)) + geom_point() + geom_smooth(method='lm')
```


## Prediction and prediction errors

Let's create a scatterplot with the least squares line laid on top.

```{r reg-with-line, eval=TRUE}
plot(mlb11$runs ~ mlb11$at_bats)
abline(m1)
```

The function `abline` plots a line based on its slope and intercept. Here, we 
used a shortcut by providing the model `m1`, which contains both parameter 
estimates. This line can be used to predict $y$ at any value of $x$. When 
predictions are made for values of $x$ that are beyond the range of the observed
data, it is referred to as *extrapolation* and is not usually recommended. 
However, predictions made within the range of the data are more reliable. 
They're also used to compute the residuals.

5.  If a team manager saw the least squares regression line and not the actual 
    data, how many runs would he or she predict for a team with 5,578 at-bats? 
    Is this an overestimate or an underestimate, and by how much? In other 
    words, what is the residual for this prediction?

```{r, echo=FALSE, results = 'asis'}
atbats <- 5578
runs.guess <- (-2789.2429 + 0.6305 * atbats) %>% round(0)
cat("Using the equation runs = -2789.2429 + 0.6305 * atbats, the team manager would predict for a team with 5,578 at-bats",runs.guess ,"runs. The residual for this prediction would be zero because this prediction was generated using the fit line. A point on the fit line has zero residuals.")
```
    

## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1) 
linearity, (2) nearly normal residuals, and (3) constant variability.

*Linearity*: You already checked if the relationship between runs and at-bats
is linear using a scatterplot. We should also verify this condition with a plot 
of the residuals vs. at-bats. Recall that any code following a *#* is intended
to be a comment that helps understand the code but is ignored by R.

```{r residuals, eval=TRUE}
plot(m1$residuals ~ mlb11$at_bats)
abline(h = 0, lty = 3)  # adds a horizontal dashed line at y = 0
```

6.  Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between runs and at-bats?

**There is no apparent pattern in the residuals plot above. They are evenly spaced aroudn the horizontal line at y = 0. This may indicate a linear relationship between the independent and response variable. **

*Nearly normal residuals*: To check this condition, we can look at a histogram

```{r hist-res, eval=TRUE}
hist(m1$residuals)
```

or a normal probability plot of the residuals.

```{r qq-res, eval=TRUE}
qqnorm(m1$residuals)
qqline(m1$residuals)  # adds diagonal line to the normal prob plot
```

7.  Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?

**The residuals are nearly normal as there are extremes at the tails of the residual histogram plot. **


*Constant variability*:

8.  Based on the plot in (1), does the constant variability condition appear to 
    be met?
```{r,fig.height=3}
ggplot(mlb11, aes(x=runs, y=at_bats)) + geom_point() + geom_smooth(method='lm')
```

**The plot in section 1 shows constants variability as we can an even distrubiotn of the data above and below the fit lines. **    

* * *

## On Your Own

-   Choose another traditional variable from `mlb11` that you think might be a 
    good predictor of `runs`. Produce a scatterplot of the two variables and fit 
    a linear model. At a glance, does there seem to be a linear relationship?

```{r,fig.height=3}
ggplot(mlb11, aes(x=runs, y=bat_avg)) + geom_point() + geom_smooth(method='lm')
```


-   How does this relationship compare to the relationship between `runs` and 
    `at_bats`? Use the R$^2$ values from the two model summaries to compare. 
    Does your variable seem to predict `runs` better than `at_bats`? How can you
    tell?

```{r m3, eval=TRUE}
m3 <- lm(runs ~ bat_avg, data = mlb11)
summary(m3)
```

Ind. Variable | R-squared  
------------- | ------------- 
`at_bats`     | 0.3729   
`bat_avg`     | 0.6561  

**The `bat_avg` is a better predictor because it explains 66% of the `runs` data variability. More than the 38% explained by the  `at_bats` variable.**

-   Now that you can summarize the linear relationship between two variables, 
    investigate the relationships between `runs` and each of the other five 
    traditional variables. Which variable best predicts `runs`? Support your 
    conclusion using the graphical and numerical methods we've discussed (for 
    the sake of conciseness, only include output for the best variable, not all 
    five).


Ind. Variable | R-squared  
------------- | ------------- 
`bat_avg`     | 0.6561
`hits`        | 0.6419
`at_bats`     | 0.3729 
`wins`        | 0.361
`strikeouts`  | 0.1694
`stolen_bases`| 0.002914


**For this section, the two variable that best predicts `runs` are `bat_avg` and `hits` . We can see in the table below how the different variables explain the variability of the data using the R-squared value. `stolen_bases` is the worst predictor as it explains less than 1% of the data variability.**


**Since by chance I chose `bat_avg` above, I will show show the second best predictor `hits` fits `runs` data. **

```{r m4, eval=TRUE}
m4 <- lm(runs ~ hits, data = mlb11)
summary(m4)
```



```{r reg-with-line@hits, eval=TRUE}
plot(mlb11$runs ~ mlb11$hits)
abline(m4)
```

-   Now examine the three newer variables. These are the statistics used by the 
    author of *Moneyball* to predict a teams success. In general, are they more 
    or less effective at predicting runs that the old variables? Explain using 
    appropriate graphical and numerical evidence. Of all ten variables we've 
    analyzed, which seems to be the best predictor of `runs`? Using the limited 
    (or not so limited) information you know about these baseball statistics, 
    does your result make sense?


Ind. Variable | R-squared  
------------- | ------------- 
`new_obs`     | 0.9349
`new_slug`    | 0.8969
`new_onbase`  | 0.8491 

**The new independent variables are a lot better predictors than the old statistics. Of interest is the stat named `obs` which I believe stands for On-base plus slugging (OPS). Using a linear regression model, this variable accounted for 93% of the data variability of `runs`.**

OPS is *..a sabermetric baseball statistic calculated as the sum of a player's on-base percentage and slugging percentage. The ability of a player both to get on base and to hit for power.*

**I does sound like a for team it would be good to have players that score high in this parameters. However, why would it be more important than the other new parameters is not so clear.**

```{r m10, eval=TRUE}
m10 <- lm(runs ~ new_obs, data = mlb11)
summary(m10)
```


```{r reg-with-line@new_obs, eval=TRUE}
plot(mlb11$runs ~ mlb11$new_obs)
abline(m10)
```





-   Check the model diagnostics for the regression model with the variable you 
    decided was the best predictor for runs.

As we can see in the plots below, the 3 conditions for the least squares line are met:
    
1. Linearity: Residuals are evenly spaced above the zeroth line. 
2. Nearly normal residuals: The residuals look normally distributed. 
3. Constant variability


```{r residuals@m10, eval=TRUE}
plot(m10$residuals ~ mlb11$at_bats)
abline(h = 0, lty = 3)  # adds a horizontal dashed line at y = 0
```

```{r hist-res@m10, eval=TRUE}
hist(m10$residuals)
```

```{r qq-res@m10, eval=TRUE}
qqnorm(m10$residuals)
qqline(m10$residuals)  # adds diagonal line to the normal prob plot
```